Niche complementarity between pollinators increases community-level plant reproductive success

### 2019-11-20 Network_structure_and_function.Rmd
affiliation
Small changes after second round of reviews in PCI Ecology

Ainhoa Magrach^1^, Francisco P. Molina^2^, Ignasi Bartomeus^2^

@@

^2Estación Biológica de Doñana (EBD-CSIC), Avda. Américo Vespucio 26, Isla de la Cartuja, 41092, Sevilla, Spain

@@

Declines in pollinator diversity and abundance have been reported across different regions, with implications for the reproductive success of plant species. However, research has focused primarily on pairwise plant-pollinator interactions, largely overlooking community-level dynamics. Yet species do not interact in isolation, they are embedded within larger networks whose structure can affect pollinator functional roles and, ultimately, the pollination services they deliver to plants. Here, we present one of the first efforts linking pollinator visitation to plant reproduction from a community-wide perspective using a well-replicated dataset encompassing 16 well-resolved plant-pollinator networks and data on reproductive success for 19 plant species from Mediterranean shrub ecosystems. We find that models including information on simple visitation metrics alone are good in explaining the variability in reproductive success observed. However, insights into the mechanisms through which differences in pollinator diversity translate into changes in reproductive success require additional information on network structure, particularly that reflecting niche complementarity between pollinators. Specifically, we find a positive effect of increasing niche complementarity between pollinators on plant reproductive success.

@@

  Indeed, theoretical research has long suggested that the structure of multitrophic communities has an effect for ecosystem functioning (reviewed in [@Thompson2012]). This line of research, rooted in niche theory and revamped by food-web studies [@Godoy2018; @Macarthur1967; @May1972a; @Tilman1982], has greatly advanced theory, but these ideas have not yet been tested using empirical data (but see [@Poisot2013]). Specifically, a major knowledge gap resides in understanding which aspects of structure determine which aspects of function [@Thompson2012]. This is because although a network perspective has promised to encapsulate complex ecological mechanisms occurring at the community level – such as indirect interactions [@Abrams1998; @Holt1977] or niche overlap [@Woodward2002]- less attention has been given to the ways in which these mechanisms relate to observed ecosystem processes [@Bluthgen2010a]. In contrast, we are now at a point in which there is considerable understanding on the attributes characterizing mutualistic interaction networks [@Bascompte2007a]. Especially, in the case of pollination, we have ample knowledge on the attributes that shape these mutualistic interactions at the community level. Amongst them is the prevalence of nested structures, i.e., arrangements where specialist species interact with a subset of the species that generalists interact with [@Bascompte2003] and which is thought to promote species diversity [@Bastolla2009]; or the relatively high extent of complementary specialization at the community scale, which may be directly related to key ecosystem functions [@Bluthgen2011]. However, the mechanisms by which these potential pathways affect plant reproduction remain to be understood [@Winfree2013]. The time is thus ripe to use the existing knowledge around plant-pollinator network structures to explore the relationship between network structure and ecosystem functioning empirically, with special emphasis being placed on the underlying ecological mechanisms that drive these relationships.

@@

At the species level, we focused on attributes defining the position of a focal plant species within the larger community. As such, we considered two metrics providing complementary non-redundant information: (i) average niche overlap in terms of pollinators between a focal plant species and each of the other plant species in the community, which estimates the potential indirect interactions between different plant species through shared resources (in this case pollinators), and (ii) centrality, which depicts the importance of the role played by a plant species within the larger community (as resource for a large number of pollinator species) and its contribution to network cohesiveness. 
 Niche overlap was calculated as the average overlap in pollinator species visiting a focal plant and each of the other plants in the community using the Morisita overlap index, a measure of similarity between two sets of data [@Zhang2016]. As a measure of centrality, we used weighted closeness centrality, which represents the number of shortest paths going through a focal plant based on a unipartite projection of the bipartite plant-pollinator network using a weighted representation of the network [@Dormann2009]. Here, links between plant species represent shared pollinator species.

@@

In order to evaluate whether adding information on network structure improves our ability to explain differences in reproductive success - both at the species and the site level - we used generalized linear (GLMs) and generalized linear mixed models (GLMMs). In both cases (species and site-level models) we fit two types of models: (i) model 1, that only included simple visitation metrics and (ii) model 2 that additionally included information on network structure. These models are meant to be additive, so that the network metrics included are intended to complement rather than substitute the simple metrics traditionally used.

  At the species level, response variables included the fruit set for different individuals of each species analyzed using a binomial distribution, the average number of seeds per fruit analyzed using a normal distribution, and the average values of fruit and seed weight fitted to Poisson distributions. The number of seeds per fruit was centered and scaled (i.e., we subtracted column means and divided by standard deviation) to allow meaningful comparisons across species with contrasting life histories. As explanatory variables, model 1 included pollinator richness, and the total number of visits received by each plant species; while model 2 added the two network attributes calculated at the species level: average plant niche overlap and centrality. For both models, we included plant species nested within site and site as random effects to account for the non-independence of several individuals measured per species and site.

@@

Figure2a<-ggplot(m2@frame,  aes(x=m2@frame$weigh_closeness.sc, y=y_partial_weigh_inv, color= Plant.sp)) +    
  geom_point(size=2.5, shape=21,color = "black", aes(fill=factor(Plant.sp))) + 
  stat_smooth(method=lm,  size=1, se=TRUE, color = "black", alpha=0.4) + 
  xlab("Centrality") +      
  ylab("Fruit set ") + theme_bw()  + theme (legend.position= "bottom") +
  theme(text = element_text(size = 18), axis.text=element_text(size=rel(1)), axis.title=element_text(size=rel(1.3), face="bold")) + 
  scale_fill_discrete(name="Plant species") +  theme(legend.text = element_text(size = 14, face = "italic"), legend.title = element_text(size=15, face = "bold"), legend.direction = "vertical") + guides(fill=guide_legend(ncol=4))

@@

m1.seed.w<-glmmTMB(mean.seedw ~  pollspsperplant.sc + tot.visits.sc  + (1|Plant.sp:Site_ID) + (1|Site_ID), family="poisson", data=s3)

@@

m2.seed.w<-glmmTMB(mean.seedw ~  pollspsperplant.sc + tot.visits.sc + weigh_closeness.sc + niche.overlap.sc  + (1|Plant.sp:Site_ID)+ (1|Site_ID), family="poisson", data=s3)

@@

m1.fruit.w<-glmmTMB(mean.fruitweight ~  pollspsperplant.sc + tot.visits.sc  + (1|Plant.sp:Site_ID)+ (1|Site_ID), family="poisson", data=s3)

@@

m2.fruit.w<-glmmTMB(mean.fruitweight ~  pollspsperplant.sc  + tot.visits.sc + weigh_closeness.sc + niche.overlap.sc  + (1|Plant.sp:Site_ID) + (1|Site_ID), family="poisson", data=s3)

@@

# extract estimates and st error for m2 and present them in paper
tableS6<-round(summary(m2.fruit.w)$coefficients$cond[,1:3], digits=2)#ch
rownames(tableS6)<-c("(Intercept)", "Pollinator richness", "Total number of visits", "Centrality", "Plant niche overlap")

```
 For all other measures of reproductive success considered (i.e., fruit and seed weight), both models had very similar fits, with ($\Delta AIC=2.2$) and 2, respectively. However, none of the variables included within our model explain the differences observed (Tables S5-S6).

@@

#save table with results for paper
tableS7<-round(summary(m1.fruitw.net)$coef[,1:3], digits=2)#ch
rownames(tableS7)<-c("(Intercept)", "Pollinator species diversity", "Total number of visits")#end of paper

@@

@@

confint(m1.seedw.net)

#save table with results for paper
tableS8<-round(summary(m1.seedw.net)$coef[,1:3], digits=2)#ch
rownames(tableS8) <-c("(Intercept)", "Pollinator species diversity", "Total number of visits")#end of paper

@@

@@

  Contrastingly, in the case of weight variables (fruit and seed weight), in both cases we find that the best model is model 1, i.e., that only including simple visitation metrics (R^2 = `r round(r2.fruitw1[2], 2)` in the case of fruit weight and `r round(r2.seedw1[2], 2)` in the case of seed weight). Here, we find a consistent positive effect of site-level pollinator richness for both weight descriptors (Tables S7-S8, Fig. 4).

@@

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, fig.height=6, fig.width=9}

@@

#######FIG S1 SAMPLING COMPLETENESS 
#FigureS2<-plot_grid( poll, plant, link, ncol = 1)
#FigureS2

@@

@@

tableS7<-round(summary(m1.fruitw.net)$coef[,1:3], digits=2)
rownames(tableS7)<-c("(Intercept)", "Pollinator richness", "Total number of visits")

kable(tableS7, caption="Table S7. Results of GLM showing effect of simple visitation metrics on site-level average fruit weight based on best model selected.")
```

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, fig.height=5}
#save table with results for paper
tableS8<-round(summary(m1.seedw.net)$coef[,1:3], digits=2)
rownames(tableS8) <-c("(Intercept)", "Pollinator richness", "Total number of visits")

kable(tableS8, caption="Table S8. Results of GLM showing effect of simple visitation metrics on site-level average seed weight based on best model selected.")
```

>>>>

Ainhoa Magrach^1^ ^2^, Francisco P. Molina^3^, Ignasi Bartomeus^3^

@@

^2^IKERBASQUE, Basque Foundation for Science, María Díaz de Haro 3, 48013, Bilbao, Spain

^3^Estación Biológica de Doñana (EBD-CSIC), Avda. Américo Vespucio 26, Isla de la Cartuja, 41092, Sevilla, Spain

@@

Declines in pollinator diversity and abundance have been reported across different regions, with implications for the reproductive success of plant species. However, research has focused primarily on pairwise plant-pollinator interactions, largely overlooking community-level dynamics. Here, we present one of the first efforts linking pollinator visitation to plant reproduction from a community-wide perspective using a well-replicated dataset encompassing 16 well-resolved plant-pollinator networks and data on reproductive success for 19 plant species from Mediterranean shrub ecosystems. We find that models including simple visitation metrics are good in explaining the variability in reproductive success observed. However, insights into the mechanisms through which differences in pollinator diversity translate into changes in reproductive success require additional information on network structure. Specifically, we find a positive effect of increasing niche complementarity between pollinators on plant reproductive success. This shows that maintaining communities with a diversity of species but also of functions is paramount to preserving natural ecosystems.

@@

  Indeed, theoretical research has long suggested that the structure of multitrophic communities has an effect for ecosystem functioning (reviewed in [@Thompson2012]). This line of research, rooted in niche theory and revamped by food-web studies [@Godoy2018; @Macarthur1967; @May1972a; @Tilman1982], has greatly advanced theory, but these ideas have not yet been tested using empirical data (but see [@Poisot2013]). Specifically, a major knowledge gap resides in understanding which aspects of structure determine which aspects of function [@Thompson2012]. This is because although a network perspective has promised to encapsulate complex ecological mechanisms occurring at the community level – such as indirect interactions [@Abrams1998; @Holt1977] or niche overlap [@Woodward2002]- less attention has been given to the ways in which these mechanisms relate to observed ecosystem processes [@Bluthgen2010a]. In contrast, we are now at a point in which there is considerable understanding on the attributes characterizing mutualistic interaction networks [@Bascompte2007a]. Especially, in the case of pollination, we have ample knowledge on the attributes that shape these mutualistic interactions at the community level. Amongst them is the prevalence of nested structures, i.e., arrangements where specialist species interact with a subset of the species that generalists interact with [@Bascompte2003] and which is thought to promote species diversity [@Bastolla2009]; or the relatively high extent of complementary specialization at the community scale, which may be directly related to key ecosystem functions [@Bluthgen2011]. However, the mechanisms by which these attributes affect plant reproduction remain to be understood [@Winfree2013]. The time is thus ripe to use the existing knowledge around plant-pollinator network structures to explore the relationship between network structure and ecosystem functioning empirically, with special emphasis being placed on the underlying ecological mechanisms that drive these relationships.

@@

At the species level, we focused on attributes defining the position of a focal plant species within the larger community. As such, we considered two metrics providing complementary non-redundant information: (i) average niche overlap in terms of pollinators between a focal plant species and each of the other plant species in the community, which estimates the potential indirect interactions between different plant species through shared resources (in this case pollinators), and (ii) centrality, which depicts the importance of the role played by a plant species within the larger community (as resource for a large number of pollinator species) and its contribution to network cohesiveness. 
 Niche overlap was calculated as the average overlap in pollinator species visiting a focal plant and each of the other plants in the community using the Morisita overlap index, a measure of similarity between two sets of data [@Zhang2016]. As a measure of centrality, we used weighted closeness centrality, which represents the number of shortest paths going through a focal plant based on a unipartite projection of the bipartite plant-pollinator network using a weighted representation of the network [@Dormann2009]. With this method, links between plant species represent shared pollinator species.

@@

In order to evaluate whether adding information on network structure improves our ability to explain differences in reproductive success - both at the species and the site level - we used generalized linear (GLMs) and generalized linear mixed models (GLMMs) respectively. In both cases (species and site-level models) we fit two types of models: (i) model 1, that only included simple visitation metrics and (ii) model 2 that additionally included information on network structure. These models are meant to be additive, so that the network metrics included are intended to complement rather than substitute the simple metrics traditionally used.

  At the species level, response variables included the fruit set for different individuals of each species analyzed using a binomial distribution and the average number of seeds per fruit, the average fruit and seed weight fitted using normal distributions. The number of seeds per fruit was centered and scaled (i.e., we subtracted column means and divided by standard deviation) to allow meaningful comparisons across species with contrasting life histories. As explanatory variables, model 1 included pollinator richness, and the total number of visits received by each plant species; while model 2 added the two network attributes calculated at the species level: average plant niche overlap and centrality. For both models, we included plant species nested within site and site as random effects to account for the non-independence of several individuals measured per species and site.

@@

Figure2a<-ggplot(m2@frame,  aes(x=m2@frame$weigh_closeness.sc, y=y_partial_weigh_inv, color= Plant.sp)) +    
  geom_point(size=2.5, shape=21,color = "black", aes(fill=factor(Plant.sp))) + 
  stat_smooth(method=glm,  method.args = list(family = "binomial"), size=1, se=TRUE, color = "black", alpha=0.4) + 
  xlab("Centrality") +      
  ylab("Fruit set ") + theme_bw()  + theme (legend.position= "bottom") +
  theme(text = element_text(size = 18), axis.text=element_text(size=rel(1)), axis.title=element_text(size=rel(1.3), face="bold")) + 
  scale_fill_discrete(name="Plant species") +  theme(legend.text = element_text(size = 14, face = "italic"), legend.title = element_text(size=15, face = "bold"), legend.direction = "vertical") + guides(fill=guide_legend(ncol=4))

@@

m1.seed.w<-glmmTMB(mean.seedw ~  pollspsperplant.sc + tot.visits.sc  + (1|Plant.sp:Site_ID) + (1|Site_ID), family="gaussian", data=s3)

@@

m2.seed.w<-glmmTMB(mean.seedw ~  pollspsperplant.sc + tot.visits.sc + weigh_closeness.sc + niche.overlap.sc  + (1|Plant.sp:Site_ID)+ (1|Site_ID), family="gaussian", data=s3)

@@

m1.fruit.w<-glmmTMB(mean.fruitweight ~  pollspsperplant.sc + tot.visits.sc  + (1|Plant.sp:Site_ID)+ (1|Site_ID), family="gaussian", data=s3)

@@

m2.fruit.w<-glmmTMB(mean.fruitweight ~  pollspsperplant.sc  + tot.visits.sc + weigh_closeness.sc + niche.overlap.sc  + (1|Plant.sp:Site_ID) + (1|Site_ID), family="gaussian", data=s3)

@@

# extract estimates and st error for m2 and present them in paper
tableS6<-round(summary(m1.fruit.w)$coefficients$cond[,1:3], digits=2)#ch
rownames(tableS6)<-c("(Intercept)", "Pollinator richness", "Total number of visits")

```
 For all other measures of reproductive success considered (i.e., fruit and seed weight), model 1 showed the best fit. However, none of the variables included within our model explain the differences observed (Tables S5-S6).

@@

#save table with results for paper
tableS7A<-round(summary(m1.fruitw.net)$coef[,1:3], digits=2)#ch
rownames(tableS7A)<-c("(Intercept)", "Pollinator species diversity", "Total number of visits")#end of paper

@@

#There is one point that has a larger pollinator diversity than the rest. To test if this point by itself might be driving the whole relationship we are repeating analyses removing this one point.

s.comm2<-subset(s.comm ,s.comm$species.poll.sc<1.9)

m1.fruitw.net2<-lm(mean.fruitweight ~  species.poll.sc + tot.visits.sc, family="gaussian", data=s.comm2)

summary(m1.fruitw.net2)

vif(m1.fruitw.net2)

m2.fruitw.net2<-lm(mean.fruitweight ~  species.poll.sc + tot.visits.sc + nodf.song.sc + functional.comp.poll.sc, family="gaussian", data=s.comm2)

summary(m2.fruitw.net2)

vif(m2.fruitw.net2)

AICtab(m1.fruitw.net2, m2.fruitw.net2)

#save table with results for paper
tableS7B<-round(summary(m1.fruitw.net2)$coef[,1:3], digits=2)#ch
rownames(tableS7B)<-c("(Intercept)", "Pollinator species diversity", "Total number of visits")#end of paper

predtestb<-summary(m1.fruitw.net2)$coefficients[1]+ (summary(m1.fruitw.net2)$coefficients[2] * s.comm2$species.poll.sc) + (summary(m1.fruitw.net2)$coefficients[3] * mean(s.comm2$tot.visits.sc))

restestb<-predtestb-s.comm2$mean.fruitweight

pltestb <- ggplot(s.comm2,  aes(x=species.poll.sc, y= restestb)) +    
  geom_point(size=2.5, shape=21, color="black", fill="darkgrey") + 
  stat_smooth(method = 'lm', size=1, se=TRUE, color = "black", alpha=0.4) +
  #ylim(0.3,0.9) +
  xlab("Pollinator richness") +      
  ylab("Fruit weight (g)") + theme_bw()  + #theme (legend.position= "none") +
  theme(text = element_text(size = 18), axis.text=element_text(size=rel(1)), axis.title=element_text(size=rel(1), face="bold"))

@@

#confint(m1.seedw.net)

#save table with results for paper
tableS8A<-round(summary(m1.seedw.net)$coef[,1:3], digits=2)#ch
rownames(tableS8A) <-c("(Intercept)", "Pollinator species diversity", "Total number of visits")#end of paper

@@

#Repeat analyses but removing one site that has a particularly large pollinator richness value to test whether this point might be driving the relationship by itself.

m1.seedw.net2<-glm(mean.seedweight ~  species.poll.sc + tot.visits.sc, family="gaussian", data=s.comm2)

summary(m1.seedw.net2)

vif(m1.seedw.net2)

m2.seedw.net2<-glm(mean.seedweight ~  species.poll.sc + tot.visits.sc + nodf.song.sc + functional.comp.poll.sc, family="gaussian", data=s.comm2)

summary(m2.seedw.net2)

vif(m2.seedw.net2)

AICtab(m1.seedw.net2, m2.seedw.net2) #m1 is better

#save table with results for paper
tableS8B<-round(summary(m1.seedw.net2)$coef[,1:3], digits=2)#ch
rownames(tableS8B) <-c("(Intercept)", "Pollinator species diversity", "Total number of visits")#end of paper

predtest2b<-summary(m1.seedw.net2)$coefficients[1]+ (summary(m1.seedw.net2)$coefficients[2] * s.comm2$species.poll.sc) + (summary(m1.seedw.net2)$coefficients[3] * mean(s.comm2$tot.visits.sc))

restest2b<-predtest2b-s.comm2$mean.seedweight

pltest2.sub <- ggplot(s.comm2,  aes(x=species.poll.sc, y= restest2b)) +    
  geom_point(size=2.5, shape=21, color="black", fill="darkgrey") + 
  stat_smooth(method = 'glm', size=1, se=TRUE, color = "black", alpha=0.4) +
  #ylim(0.3,0.9) +
  xlab("Pollinator richness") +      
  ylab("Seed weight (g)") + theme_bw()  + #theme (legend.position= "none") +
  theme(text = element_text(size = 18), axis.text=element_text(size=rel(1)), axis.title=element_text(size=rel(1), face="bold"))

FigureS3<-plot_grid(pltestb, pltest2.sub, labels = c("A", "B"), nrow=1)

@@

  Contrastingly, in the case of weight variables (fruit and seed weight), in both cases we find that the best model is model 1, i.e., that only including simple visitation metrics (R^2 = `r round(r2.fruitw1[2], 2)` in the case of fruit weight and `r round(r2.seedw1[2], 2)` in the case of seed weight). Here, we find a consistent positive effect of site-level pollinator richness for both weight descriptors (Tables S7A-S8A, Fig. 4). This effect is maintained even after removing a site that has a particularly large pollinator richness value (Tables S7B-S8B, Fig. S3).

@@

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, fig.height=9, fig.width=9}

@@

#######FIG S2 SAMPLING COMPLETENESS 
#FigureS2<-plot_grid( poll, plant, link, ncol = 1)
FigureS2

@@

FigureS3

**Figure S3.** Figure S3. Partial residual plots showing the effect of pollinator richness on site-level average A) fruit and B) seed weight. Here, a site with a particularly large pollinator richness value is removed to test whether it might be driving the significant relationship. Dots represent values for each site (N=15 sites).

@@

tableS7A<-round(summary(m1.fruitw.net)$coef[,1:3], digits=2)
rownames(tableS7)<-c("(Intercept)", "Pollinator richness", "Total number of visits")

kable(tableS7A, caption="Table S7. Results of GLM showing effect of simple visitation metrics on A) site-level average fruit weight based on best model selected and B) the same analysis removing one site that has a particularly large pollinator richness value to test whether this point might be driving the relationship.")
```

tableS7B<-round(summary(m1.fruitw.net2)$coef[,1:3], digits=2)
rownames(tableS7B)<-c("(Intercept)", "Pollinator richness", "Total number of visits")

kable(tableS7B, caption="Table S7. Results of GLM showing effect of simple visitation metrics on A) site-level average fruit weight based on best model selected and B) the same analysis removing one site that has a particularly large pollinator richness value to test whether this point might be driving the relationship.")
```

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, fig.height=5}
#save table with results for paper
tableS8A<-round(summary(m1.seedw.net)$coef[,1:3], digits=2)
rownames(tableS8A) <-c("(Intercept)", "Pollinator richness", "Total number of visits")

kable(tableS8A, caption="Results of GLM showing effect of simple visitation metrics on A) site-level average seed weight based on best model selected and B) the same analysis removing one site that has a particularly large pollinator richness value to test whether this point might be driving the relationship.")
```

```{r, include=TRUE, echo = FALSE, message=FALSE, error=FALSE, warning=FALSE, fig.height=5}
#save table with results for paper
tableS8B<-round(summary(m1.seedw.net2)$coef[,1:3], digits=2)
rownames(tableS8B) <-c("(Intercept)", "Pollinator richness", "Total number of visits")

kable(tableS8B, caption="Results of GLM showing effect of simple visitation metrics on A) site-level average seed weight based on best model selected and B) the same analysis removing one site that has a particularly large pollinator richness value to test whether this point might be driving the relationship.")
```
